---
title: 'Reproducible Research: Peer Assessment 1'
output:
  pdf_document: default
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data

_Unzip the data into "activity.csv" file and Load the data into "activity" dataFrame See the dimensions, head, structure of the activity dataFrame_

```{r}
if(!file.exists("activity.csv")) {
  unzip("reproducible_research_week2_project/RepData_PeerAssessment1/activity.zip", exdir = "reproducible_research_week2_project/RepData_PeerAssessment1")
}
activity = read.csv("activity.csv")
```

```{r}
dim(activity)
```

```{r}
head(activity)
```

```{r}
str(activity)
```



_Transforming the class of activity$date to Date format_

```{r}
activity$date = as.Date(as.character(activity$date), "%Y-%m-%d")
str(activity)
```


## What is mean total number of steps taken per day?


_Lets calculate total number of steps taken per day using tapply excluding NAs and see some values using head_

```{r}
total_numberofsteps_perday = with(activity, tapply(steps, date, sum, na.rm = TRUE))
head(total_numberofsteps_perday)
```

_Lets make a histogram of the total number of steps taken each day_
```{r}
hist(total_numberofsteps_perday,breaks = 20)
```

_Lets calculate and report the mean and median of the total number of steps taken per day_

```{r}
mean(total_numberofsteps_perday)
```


```{r}
median(total_numberofsteps_perday)
```


## What is the average daily activity pattern?


_Lets make a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)_

_Lets store x-axis data in fivemin_interval variable using unique function on activity$interval_
```{r}
fivemin_interval = unique(activity$interval)
str(fivemin_interval)
```

_And lets store y-axis data as below_
```{r}
averagenumberofstepstaken_averagedacrossalldays = with(activity,tapply(steps, interval, mean, na.rm=TRUE))
str(averagenumberofstepstaken_averagedacrossalldays)
```

_And here comes the plot_

```{r}
#png(filename = "instructions_fig/plot1.png")
plot(fivemin_interval,averagenumberofstepstaken_averagedacrossalldays,
     type = "l", xlab = "Interval" , ylab = "Number of steps", main = "Average daily activity pattern" )
#dev.off()
```

_Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?_

Lets print averagenumberofstepstaken_averagedacrossalldays in ascending order and get the index of its maximum.
```{r}
order(averagenumberofstepstaken_averagedacrossalldays)
```

_Its largest value is in index 104. Lets see the interval and maximum average steps_

```{r}
averagenumberofstepstaken_averagedacrossalldays[104]
```


## Imputing missing values


_Lets calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)_

```{r}
sum(is.na(activity$steps))
```

_Lets devise a strategy for filling in all of the missing values in the dataset. Here lets impute missing value with mean of that interval and lets create a new dataset (activity_impute) that is equal to the original dataset but with the missing data filled in._

```{r}
activity_imputed= activity
missingData = is.na(activity$steps)
meanValuesByInterval = tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
activity_imputed$steps[missingData] =  meanValuesByInterval[as.character(activity_imputed$interval[missingData])]
sum(is.na(activity_imputed))
```

_Lets make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day._

```{r}
total_numberofsteps_perday_imputed = with(activity_imputed, tapply(steps, date, sum))
hist(total_numberofsteps_perday_imputed, breaks = 20)
mean(total_numberofsteps_perday_imputed)
median(total_numberofsteps_perday_imputed)
```

_Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?_

```{r}
mean(total_numberofsteps_perday)
mean(total_numberofsteps_perday_imputed)
```

Yes, Mean differs

```{r}
median(total_numberofsteps_perday)
median(total_numberofsteps_perday_imputed)
```

Yes, Median differs


## Are there differences in activity patterns between weekdays and weekends? Lets use dataset (activity_imputed) with filled-in missing values 

_Lets create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day._

```{r}
library(chron)
week_factor = is.weekend(activity_imputed$interval)
activity_imputed_weekend = activity_imputed[week_factor,]
activity_imputed_weekday = activity_imputed[!week_factor,]
average_numberofsteps_perday_imputed_weekday = with(activity_imputed_weekday, tapply(steps, interval, mean))
average_numberofsteps_perday_imputed_weekend = with(activity_imputed_weekend, tapply(steps, interval, mean))
```

_ Lets make a panel plot containing a time series plot (i.e.type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)._

```{r}
par(mfrow = c(2,1))
plot(unique(activity_imputed_weekday$interval) , average_numberofsteps_perday_imputed_weekday,
     type="l" , xlab = "Interval",ylab = "Frequency", main = " weekday days")
plot(unique(activity_imputed_weekend$interval) , average_numberofsteps_perday_imputed_weekend,
     type="l", xlab = "Interval",ylab = "Frequency", main = " weekend days")

```

From the plots we can assumer that there is the hike in steps beginning from the start of the day in weekdays
and in weekends the average is shared equally among the intervals compared to the weekdays





